---
title: "Project Showcase Figure"
author: "Rachel Gonsenhauser"
geometry: margin=2.54cm
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
  word_document: default
mainfont: Times New Roman
subtitle: 
fontsize: 12pt
---

```{r setup, include=FALSE}
# Setting up session
getwd()
library(dplyr)
library(tidyverse)
library(lubridate)

# Pulling in processed dataset
Water.Quality.And.Income.Processed <- read.csv("../Data/Processed/CDC_WaterQualityAndincome_Processed.csv")

mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

Water.Quality.And.Income.Processed$Year <- as.character(Water.Quality.And.Income.Processed$Year) 
Water.Quality.And.Income.Processed$State <- as.character(Water.Quality.And.Income.Processed$State) 
Water.Quality.And.Income.Processed$Arsenic_ugL <- as.numeric(Water.Quality.And.Income.Processed$Arsenic_ugL) 

# Southeast data 
Southeast.States.data <- 
  filter(Water.Quality.And.Income.Processed, State == "North Carolina" | State == "South Carolina" | State == "Florida" | State == "Georgia" | State == "Virginia" | State == "Alabama" | State == "Tennessee" | State == "West Virginia" | State == "Mississippi" | State == "Kentucky" | State == "Louisiana") 

```

```{r figs, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap="Mean annual arsenic concentration by size of population served by community water system (CWS) in southeastern states."}

Southeast.arsenic.plot <- ggplot(Southeast.States.data, 
                 aes(x = Population.Served, y = Arsenic_ugL)) +
   geom_hline(yintercept = 10, lty = 1) +
   geom_text(x = 2000000, y = 150, label = "Maximum contaminant level", hjust = 1) +
  geom_point(aes(color = State)) +
  xlim(0, 2300000) +
     scale_color_brewer(palette = "Paired") +
   labs(x = expression("Population served"), y = expression(Arsenic ~ (mu*g / L)),
       color = "State", shape = "") + 
    theme(legend.position = "bottom",
       legend.text = element_text(size = 10), legend.title = element_text(size = 10))
print(Southeast.arsenic.plot)
```

>There is a lot of variation in the mean annual concentration of arsenic in drinking water detected in small CWSs. As the amount of people served by a water system increases, the concentration of arsenic in water is less variable and is closer to the maximum contaminant level (MCL) threshold. It seems that this trend holds for most southeastern states and that Florida, North Carolina, and Kentucky are the only states with water systems that served over 1 million people (Figure 1).